---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file, then run
     devtools::build_readme() and commit both files. -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# BREAD

<!-- badges: start -->
[![R-CMD-check](https://github.com/BacZemin/BREAD/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/BacZemin/BREAD/actions/workflows/R-CMD-check.yaml)
[![bioc-check](https://github.com/BacZemin/BREAD/actions/workflows/bioc-check.yaml/badge.svg)](https://github.com/BacZemin/BREAD/actions/workflows/bioc-check.yaml)
[![pkgdown](https://github.com/BacZemin/BREAD/actions/workflows/pkgdown.yaml/badge.svg)](https://github.com/BacZemin/BREAD/actions/workflows/pkgdown.yaml)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![License: MIT](https://img.shields.io/badge/license-MIT-blue.svg)](https://opensource.org/licenses/MIT)
[![Docs](https://img.shields.io/badge/docs-pkgdown-blue.svg)](https://baczemin.github.io/BREAD/)
<!-- badges: end -->

**BREAD** (Bayesian Region-specific DNA methylation inference) provides
targeted Bayesian inference for **predefined** DNA methylation regions from
array data stored in `SummarizedExperiment` objects. For each region you
supply, BREAD fits a Bayesian model, computes the posterior probability of a
directional methylation change, and classifies the region as
**hypermethylated**, **hypomethylated**, or **inconclusive** at
user-configurable effect-size and probability thresholds.

Unlike genome-wide DMR callers that scan for regions, BREAD answers a
different question: *given regions I already care about (PRC2 targets, CGIs,
LADs, a chromHMM state, a custom BED), what is the posterior evidence for
methylation change in each one, and how confident am I?* Output is a
per-region posterior — effect size, credible interval, and directional
probabilities — not just a p-value.

## Installation

BREAD is in development. Install the latest version from GitHub:

``` r
# install.packages("pak")
pak::pak("BacZemin/BREAD")

# or with remotes:
# install.packages("remotes")
remotes::install_github("BacZemin/BREAD")
```

BREAD depends on Bioconductor packages (`SummarizedExperiment`,
`GenomicRanges`, `S4Vectors`). The optional `brms` backend additionally needs
`brms` plus a working Stan toolchain.

## Example

BREAD ships a small packaged dataset so you can run the whole pipeline out of
the box: 8 EPICv2 arrays from a fibroblast passage-aging × vitamin C
experiment, plus 500 predefined regions spanning five feature classes
(PMD, PRC-CGI, bivalent, ...).

```{r example, message = FALSE}
library(BREAD)
suppressPackageStartupMessages(library(SummarizedExperiment))

se  <- readRDS(system.file("extdata", "vitc_ag06561.rds", package = "BREAD"))
reg <- readRDS(system.file("extdata", "vitc_regions.rds", package = "BREAD"))

# Which regions change methylation with passage in the control fibroblasts?
se_ctrl <- se[, se$condition == "ctrl"]

fit <- fit_bread(
  se_ctrl,
  features          = reg,
  design            = ~ passage,
  feature_class_col = "feature_class"
)

fit                     # summary: n regions, backend, classification counts
```

```{r example-results}
res <- results(fit)     # one row per region
table(res$classification)
```

That is the whole pattern: a `SummarizedExperiment` of array data, a
`GRanges` of regions to test, and a model formula referencing columns of
`colData(se)`. `fit_bread()` auto-detects the assay and whether values are
on the beta or M scale, so the three arguments above are usually all you
need. `feature_class_col` is optional — supply it when your regions carry a
grouping column you want summarized.

### Reading the results

`results(fit)` returns one row per region:

| column | meaning |
|---|---|
| `region_id` | region identifier |
| `n` | number of probes summarized in the region |
| `mean_effect` | posterior mean methylation change (M-scale) |
| `ci_lo`, `ci_hi` | 95% credible interval |
| `prob_hyper` | P(effect > +delta) — evidence for hypermethylation |
| `prob_hypo` | P(effect < -delta) — evidence for hypomethylation |
| `classification` | `hypermethylated` / `hypomethylated` / `inconclusive` |

```{r results-head}
head(res[, c("region_id", "n", "mean_effect", "ci_lo", "ci_hi",
             "classification")])
```

A region is called **hypermethylated** if `prob_hyper >= prob_cutoff`,
**hypomethylated** if `prob_hypo >= prob_cutoff`, otherwise
**inconclusive**. Defaults are `delta = 0.10` (M-scale) and
`prob_cutoff = 0.95`; both are arguments to `fit_bread()`.

`classifications(fit)` returns just the per-region calls, and
`posterior_draws(fit)` gives posterior samples for downstream summaries.

### Backends

- `backend = "conjugate"` (default) — analytic Normal-Inverse-Gamma
  posterior, no MCMC. Hundreds of regions fit in well under a second.
- `backend = "brms"` — full MCMC via Stan; compiles once, then reuses the
  compiled model across regions. Use when you need the flexibility of a
  full Bayesian fit.

### KnowYourCG enrichment

`bread_kycg()` takes the probes in your hyper- or hypo-classified regions
and runs `knowYourCG::testEnrichment()` against curated CpG databases, so
you can ask what genomic features your called regions are enriched for.

## Vignettes

Two worked examples on real data (rendered on the
[documentation site](https://baczemin.github.io/BREAD/)):

- **Getting started** (`bread-intro`) — TCGA HM450 matched normal/tumour
  pairs over chromHMM chromatin-state regions, end-to-end through
  `bread_kycg()`.
- **Vitamin C EPICv2** (`bread-vitc`) — the packaged fibroblast
  passage-aging × vitamin C experiment, recovering classic PMD-hypo /
  PRC-CGI-hyper region signatures.

## Status

Milestone 1 (MVP) is complete: the full `fit_bread()` pipeline, conjugate
and brms backends, S4 classes with accessors, plotting helpers, KYCG
integration, two real-data vignettes, and a live pkgdown site. Partial
pooling across regions, feature-class priors, random-effects designs, and
contrast vectors are on the roadmap. See `NEWS.md` for the changelog.

## Citation

BREAD does not have an associated publication yet. For now, cite the
package itself:

```{r citation, comment = ""}
citation("BREAD")
```

## License

MIT © Jaemin Park. See `LICENSE`.
