---
title: "Intro to BREAD"
subtitle: "Bayesian region-specific DNA methylation inference"
author: "Jaemin Park"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Intro to BREAD}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.align = "center", fig.width = 7, fig.height = 4.5, dpi = 120
)
set.seed(2026)
```

## Why BREAD?

Many methylation studies are not purely discovery-oriented. Instead of
asking *"which CpGs are significant genome-wide?"*, the question is
often region-centric:

- Are **bivalent promoters** hypermethylated in tumor vs. normal?
- Do **PMD-like heterochromatin regions** lose methylation in aging?
- Which of my **predefined chromatin classes** show coherent change?

BREAD answers those questions directly, returning posterior
probabilities per region at a biologically meaningful effect-size
threshold.

## A real-data demo: TCGA matched tumor–normal (HM450)

We use `HM450.76.TCGA.matched` from **`sesameData`** — 10,042 HM450 probes
× 76 samples (38 patients each with matched Normal + Tumour). This is a
strong, well-understood signal: tumors show broad hypermethylation at
Polycomb targets and hypomethylation in partially-methylated / quiescent
regions.

```{r load, message = FALSE}
library(BREAD)
library(sesameData)
library(knowYourCG)
library(SummarizedExperiment)
library(GenomicRanges)
library(ggplot2)
library(patchwork)

tcga      <- sesameData::sesameDataGet("HM450.76.TCGA.matched")
probeInfo <- sesameData::sesameDataGet("HM450.probeInfo")
```

```{r cache, include = FALSE}
sesameData::sesameDataCache(c(
  "HM450.76.TCGA.matched",
  "HM450.probeInfo"
))
```

```{r make-se}
probes <- probeInfo$mapped.probes.hg38
# keep only probes present in the TCGA beta matrix
keep   <- intersect(rownames(tcga$betas), names(probes))
se <- SummarizedExperiment(
  assays    = list(betas = tcga$betas[keep, , drop = FALSE]),
  rowRanges = probes[keep],
  colData   = tcga$sampleInfo
)
se$type <- factor(se$type, levels = c("Normal", "Tumour"))
se
```

## Define region sets via KnowYourCG

We use KnowYourCG's HM450 chromHMM database to group probes by
chromatin state, then merge nearby same-state probes into intervals.
This gives us biologically meaningful feature classes with no manual
annotation work.

```{r regions}
chromhmm <- knowYourCG::getDBs("KYCG.HM450.chromHMM.20211020")
# chromhmm is a named list: state -> character vector of probe IDs
str(chromhmm, list.len = 3, give.attr = FALSE)

# Keep a few interpretable classes
state_map <- c(
  BivProm   = "10_TssBiv",
  ReprPC    = "13_ReprPC",
  ActiveTSS = "1_TssA",
  Enh       = "7_Enh"
)
state_probes <- chromhmm[names(chromhmm) %in% state_map]
# Rename to friendly labels
names(state_probes) <- names(state_map)[match(names(state_probes), state_map)]

# Build regions per class: intersect with our SE probes, merge by ±2 kb
features <- do.call(c, unname(lapply(names(state_probes), function(cls) {
  pids <- intersect(state_probes[[cls]], rownames(se))
  if (length(pids) < 3L) return(GRanges())
  gr  <- rowRanges(se)[pids]
  red <- GenomicRanges::reduce(GenomicRanges::resize(gr, 4001L, fix = "center"))
  ov  <- GenomicRanges::countOverlaps(red, gr)
  red <- red[ov >= 3L]
  # Cap per class for speed
  if (length(red) > 120L) red <- sort(red[sample.int(length(red), 120L)])
  mcols(red)$feature_class <- cls
  names(red) <- sprintf("%s_%03d", cls, seq_along(red))
  red
})))
table(features$feature_class)
```

## Fit BREAD with one call

`fit_bread()` auto-detects the assay (`"betas"`) and the scale (`"Beta"`
from the value range), so the minimal call is:

```{r fit}
fit <- fit_bread(se, features, design = ~ type,
                 feature_class_col = "feature_class")
fit
```

Under the hood BREAD is running an analytic conjugate posterior per
region; 480 regions fit in well under a second on a laptop. Default
thresholds (`delta = 0.10`, `prob_cutoff = 0.95`) can be overridden if
you want to be stricter or looser.

## Results

```{r results}
res <- results(fit)
head(res[order(res$p_gt_delta, decreasing = TRUE),
         c("region_id", "mean_effect", "ci_lo", "ci_hi",
           "p_gt_delta", "p_lt_neg_delta", "classification")], 5)
```

```{r by-class}
mp <- unique(fit@mapping[, c("region_id", "feature_class")])
cls_tbl <- merge(res[, c("region_id", "classification")], mp,
                 by = "region_id")
table(cls_tbl$feature_class, cls_tbl$classification)
```

Classic tumor methylome signatures come through:

- **Polycomb-repressed (ReprPC)** and **bivalent promoters (BivProm)**
  hypermethylate — the classic TCGA "tumor CpG island methylator
  phenotype"-adjacent signal.
- **Active TSS** largely unchanged at region level.

## Visualize

### Classification summary

```{r feature-set, fig.height = 3.5}
plot_feature_set(fit, feature_class_col = "feature_class") +
  ggtitle("Region classifications by chromatin class")
```

### One region: posterior density + raw beta

Pick the top hypermethylated region and inspect the posterior +
underlying values. The x-axis preserves the factor order we set on
`se$type` — Normal on the left, Tumour on the right.

```{r one-region, fig.width = 10, fig.height = 4}
top_hyper <- res[res$classification == "hypermethylated", ]
top_hyper <- top_hyper[order(top_hyper$p_gt_delta, decreasing = TRUE), ]
rid <- top_hyper$region_id[1]

p1 <- plot_region_posterior(fit, region_id = rid) +
  ggtitle(sprintf("Posterior: %s", rid))
p2 <- plot_region_data(fit, region_id = rid) +
  ggtitle("Raw region means by group")
p1 | p2
```

### Posterior draws

```{r draws, fig.height = 3.5}
d <- posterior_draws(fit, region_id = rid, n = 2000L, seed = 1L)
ggplot(d, aes(x = value)) +
  geom_histogram(
    bins = 40,
    fill = bread_colors("classification")[["hypermethylated"]],
    alpha = 0.85
  ) +
  geom_vline(
    xintercept = c(-0.10, 0, 0.10),
    linetype   = c("dashed","dotted","dashed"),
    color      = c("gray60","gray40","gray60")
  ) +
  labs(x = "effect (M-value units)", y = "count",
       title = sprintf("Posterior draws for %s", rid)) +
  theme_classic(base_size = 12)
```

## Enrichment: what's driving the hyper/hypo sets?

`bread_kycg()` hands the probes in hyper- or hypo-classified regions
off to `knowYourCG::testEnrichment()` against curated HM450 databases
(TFBS, chromHMM, tissue signatures, ...). The universe is the set of
probes BREAD actually tested.

```{r kycg, eval = requireNamespace("knowYourCG", quietly = TRUE), message = FALSE}
enr <- bread_kycg(
  fit,
  which    = c("hypermethylated", "hypomethylated"),
  platform = "HM450",
  databases = c(
    "KYCG.HM450.chromHMM.20211020",
    "KYCG.HM450.TFBSconsensus.20211013"
  )
)
head(enr[order(enr$FDR), c("query", "dbname", "estimate", "p.value", "FDR")], 10)
```

The top hits on the hyper set should be Polycomb / bivalent-promoter
states and Polycomb-complex TFs (SUZ12, EZH2, BMI1) — the TCGA
tumor-methylome signature.

## What BREAD is doing

For each region *r*, the default backend fits a conjugate
Normal–Inverse-Gamma regression on the region-level mean beta
(transformed to M):

$$
y_{ir} \sim \mathcal{N}(X_i\beta_r,\ \sigma_r^2),\quad
\beta_r\mid\sigma_r^2 \sim \mathcal{N}(\mu_0,\ \sigma_r^2\Lambda_0^{-1}),\quad
\sigma_r^2 \sim \text{Inv-Gamma}(a_0, b_0).
$$

The marginal posterior of the contrast coefficient is a location–scale
Student-t with `df = 2 a_n`. BREAD uses `pt()`/`qt()` to compute
$P(\beta > \delta)$ and $P(\beta < -\delta)$ analytically — no MCMC.

The classification rule is then:

- **hypermethylated** if $P(\beta > \delta) \ge$ `prob_cutoff`,
- **hypomethylated** if $P(\beta < -\delta) \ge$ `prob_cutoff`,
- **inconclusive** otherwise.

Need partial pooling across regions, non-conjugate priors, or ordered
contrasts? Set `backend = "brms"`; everything else stays the same.

## Session info

```{r session}
sessionInfo()
```
