---
title: "BREAD on real EPICv2 data: vitamin C in serially passaged fibroblasts"
author: "Jaemin Park"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{BREAD on real EPICv2 data: vitamin C in serially passaged fibroblasts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.align = "center", fig.width = 7, fig.height = 4.5, dpi = 120
)
set.seed(2026)
```

```{r logo, echo=FALSE, results="asis"}
.logo <- "../man/figures/logo.png"
if (file.exists(.logo)) {
  cat(sprintf(
    '<img src="%s" align="right" width="115" alt="BREAD hex logo" style="margin: 0 0 1em 1em;" />',
    knitr::image_uri(.logo)
  ))
}
```

## Biological question

Ascorbic acid (vitamin C) is a cofactor for TET dioxygenases, which oxidize
5mC → 5hmC → 5fC → 5caC and drive active DNA demethylation. TET activity is
thought to protect CpG islands from aberrant age-associated hypermethylation.
In serially passaged human fibroblasts, two CpG sets consistently change:

- **PRC-CGI** (Polycomb-marked CpG islands, H3K27me3 ∩ CGI) **gain**
  methylation with passage number;
- **PMD-soloWCGW** (late-replicating solo-WCGW CpGs within partially
  methylated domains) **lose** methylation with passage number.

The question the experiment asks: *does VitC supplementation attenuate
either signal?* A conventional probe-level DMP pipeline reports significant
CpGs; BREAD asks the complementary, region-centric question directly:

> For each predefined feature region, what is the posterior probability
> that the methylation effect exceeds a biologically meaningful threshold?

## The packaged data

BREAD ships a small (~0.5 MB) subset of a real EPICv2 experiment from our
lab — 8 arrays on AG06561 fetal skin fibroblasts across a 2 × 2 design
(condition × passage × 2 technical replicates). The data-raw builder is
`data-raw/make_vitc_extdata.R` (in the source repository).

```{r load-data, message = FALSE}
library(BREAD)
library(SummarizedExperiment)
library(GenomicRanges)
library(ggplot2)
library(patchwork)

se  <- readRDS(system.file("extdata", "vitc_ag06561.rds", package = "BREAD"))
reg <- readRDS(system.file("extdata", "vitc_regions.rds", package = "BREAD"))
se
```

```{r colData}
as.data.frame(colData(se)[, c("condition", "passage", "replicate")])
```

```{r regions}
length(reg); table(reg$feature_class)
```

Five feature classes, 100 regions each, all on autosomes, each region
spanning ≥ 3 EPICv2 probes after a ±2 kb merge. `reg$feature_class` will
drive the `plot_feature_set()` stratification below.

## Aging contrast in ctrl samples

Under the `ctrl` condition (no VitC), BREAD should recover the well-known
aging signatures: PRC-CGI hypermethylation and PMD-soloWCGW
hypomethylation.

```{r fit-aging}
se_ctrl <- se[, colData(se)$condition == "ctrl"]
fit_aging <- fit_bread(
  se            = se_ctrl,
  features      = reg,
  design        = ~ passage,
  contrast      = "passagelate",
  delta         = 0.10,
  prob_cutoff   = 0.95,
  min_probes    = 3L,
  feature_class_col = "feature_class"
)
fit_aging
```

### Classifications by class

```{r aging-by-class}
mp       <- unique(fit_aging@mapping[, c("region_id", "feature_class")])
res_a    <- merge(results(fit_aging)[, c("region_id", "classification")],
                  mp, by = "region_id")
table(res_a$feature_class, res_a$classification)
```

As expected:
- **PRC-CGI and Bivalent** regions skew strongly toward `hypermethylated`
  with passage.
- **PMD-soloWCGW** regions skew strongly toward `hypomethylated`.
- **Active promoters** and **active enhancers** are largely `inconclusive`
  with this sample size (n = 2 per passage), consistent with the weaker
  direct effect at those regulatory elements.

### Bird's-eye plot

```{r aging-feature-set, fig.height = 3.5}
plot_feature_set(fit_aging, feature_class_col = "feature_class") +
  ggtitle("Aging (late vs early), ctrl samples")
```

## VitC contrast at late passage

Now condition on late passage and test whether ascorbic acid
supplementation demethylates any regions.

```{r fit-vitc}
se_late <- se[, colData(se)$passage == "late"]
fit_vitc <- fit_bread(
  se            = se_late,
  features      = reg,
  design        = ~ condition,
  contrast      = "conditionaa57",
  delta         = 0.10,
  prob_cutoff   = 0.95,
  min_probes    = 3L,
  feature_class_col = "feature_class"
)
fit_vitc
```

```{r vitc-by-class}
res_v <- merge(results(fit_vitc)[, c("region_id", "classification")],
               mp, by = "region_id")
table(res_v$feature_class, res_v$classification)
```

```{r vitc-feature-set, fig.height = 3.5}
plot_feature_set(fit_vitc, feature_class_col = "feature_class") +
  ggtitle("VitC (aa57 vs ctrl), late passage")
```

VitC demethylation is broadly distributed but, with only two replicates per
arm, most regions land in the `inconclusive` class at `prob_cutoff = 0.95`.

That word now carries a precise meaning. With n = 2 the posterior is wide,
so almost nothing reaches `prob_rope >= 0.95` either: these regions are not
being called *unchanged*, they are being called *unresolved*. BREAD is
saying it can neither detect a $\delta = 0.10$ effect nor rule one out —
which is the honest answer, and a strictly more informative one than a
non-significant p-value, because the same table tells you which regions
*did* resolve in each direction.

```{r rope-at-n2}
summary(results(fit_vitc)$prob_rope)
```

If most of that distribution sits well below 0.95, the experiment is
underpowered rather than null. Contrast this with a well-powered design,
where regions genuinely unaffected by the treatment accumulate `prob_rope`
near 1 and get called `unchanged` — a positive claim of no effect.

## The biologically interesting intersection

The most interpretable cross-tab is *aging × VitC*: which regions that
hypermethylate with aging are demethylated by vitamin C?

```{r cross-tab}
a <- classifications(fit_aging)
v <- classifications(fit_vitc)
common <- intersect(names(a), names(v))
xt <- table(aging = a[common], vitc = v[common])
xt
```

The **aging-hyper ∩ VitC-hypo** cell contains the candidate
*VitC-responsive aging hypermethylation* regions — the mechanistic story the
project is chasing.

```{r vitc-protected}
protected <- names(a)[a == "hypermethylated" & v == "hypomethylated"]
length(protected)
# Feature-class composition of the protected set
table(mp$feature_class[mp$region_id %in% protected])
```

The protected set is enriched in **PRC-CGI / Bivalent** regions — consistent
with vitamin C's known role as a CpG-island guardian via TET recruitment.

## Zooming in on one region

```{r protected-first, fig.height = 3.8}
if (length(protected) > 0L) {
  rid <- protected[1]
  rid
  plot_region_posterior(fit_aging, region_id = rid) +
    ggtitle(sprintf("Aging posterior for %s", rid))
}
```

```{r protected-data, fig.width = 9, fig.height = 4}
if (length(protected) > 0L) {
  rid <- protected[1]
  p_aging <- plot_region_data(fit_aging, rid) +
    ggtitle("Aging: late vs early (ctrl)")
  p_vitc  <- plot_region_data(fit_vitc,  rid) +
    ggtitle("VitC: aa57 vs ctrl (late)")
  p_aging + p_vitc
}
```

```{r draws, fig.height = 3.5}
if (length(protected) > 0L) {
  rid <- protected[1]
  d <- posterior_draws(fit_vitc, region_id = rid, n = 2000L, seed = 1L)
  ggplot(d, aes(x = value)) +
    geom_histogram(fill = bread_colors("classification")[["hypomethylated"]],
                   bins = 40, alpha = 0.85) +
    geom_vline(xintercept = c(-0.10, 0, 0.10),
               linetype = c("dashed","dotted","dashed"),
               color = c("gray60","gray40","gray60")) +
    labs(title = sprintf("VitC posterior for %s", rid),
         x = "effect (M-value units)", y = "count") +
    theme_classic(base_size = 12)
}
```

## Caveats

- **n = 2 per arm.** The credible intervals are wide and many regions
  remain `inconclusive` — genuinely unresolved, not shown to be flat (see
  the `prob_rope` distribution above). BREAD's posterior probabilities are
  the honest answer given the data; lowering `prob_cutoff` to, say, `0.80`
  will promote more regions to `hyper` / `hypo` but also admit more false
  positives. Use `refit_bread(fit, prob_cutoff = 0.80)` to sweep that
  without re-fitting anything.
- Technical replicates, not biological. Real variance in the `aa57`
  effect across fibroblast lines is not captured by this experiment and
  would produce additional dispersion if included.
- Feature-class regions here are constructed from the EPICv2 manifest's
  probe-level annotations (CGI, PMD, histone marks) by merging nearby
  annotated probes. Users should consider higher-resolution external
  references (ENCODE chromHMM, segway, compartment-aware WGBS-derived PMD
  calls) when the question demands it.

## Session info

```{r session}
sessionInfo()
```
