---
title: "Intro to BREAD"
subtitle: "Bayesian region-specific DNA methylation inference"
author: "Jaemin Park"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Intro to BREAD}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.align = "center",
  fig.width  = 7,
  fig.height = 4.5,
  dpi = 120
)
set.seed(2026)
```

## What BREAD is for

Many methylation studies are not purely discovery-oriented. Instead of asking
*"which CpGs are significant genome-wide?"*, the question is often:

- Are **PRC2 target regions** hypermethylated in old vs. young samples?
- Do **bivalent promoters** coherently lose methylation after treatment?
- Among my **predefined chromatin classes** (LADs, PMDs, CGIs, …), which
  show strong directional evidence at a biologically meaningful effect size?

Existing tooling is strong for genome-wide hypothesis testing, but answering
those questions directly benefits from:

1. **region-centric** inference (one answer per region, not per CpG),
2. **threshold-based** decisions on biologically meaningful effect sizes,
3. **posterior probabilities** rather than p-values,
4. *borrowing strength across probes inside a region.*

BREAD provides a minimal, interpretable Bayesian workflow for exactly these
targeted questions, built around `SummarizedExperiment` and `GRanges`.

## Install and load

```{r load, message = FALSE}
library(BREAD)
library(SummarizedExperiment)
library(GenomicRanges)
library(ggplot2)
library(patchwork)
```

## Build a toy young-vs-old methylation experiment

We simulate 60 CpG probes on chromosome 1 spaced every 1 kb, across
16 samples (8 young, 8 old). We define five regions of interest — a mix of
PRC-like, CGI-like, and LAD-like feature classes — and inject a strong hyper
or hypo signal into a few of them so we can see BREAD recover them.

```{r toy-data}
n_probes  <- 60L
n_samples <- 16L

# Probe-level GRanges (rowRanges of the SE)
probes <- GRanges(
  seqnames = "chr1",
  ranges   = IRanges(start = seq(1L, by = 1000L, length.out = n_probes),
                     width = 1L)
)
names(probes) <- sprintf("cg%04d", seq_len(n_probes))

# Sample metadata
coldata <- S4Vectors::DataFrame(
  group = factor(rep(c("young", "old"), each = n_samples / 2),
                 levels = c("young", "old")),
  sex   = factor(rep(c("F", "M"), length.out = n_samples)),
  row.names = sprintf("S%02d", seq_len(n_samples))
)

# Baseline M-values: noise centered at 0
X <- matrix(
  rnorm(n_probes * n_samples, mean = 0, sd = 0.3),
  nrow = n_probes,
  dimnames = list(names(probes), rownames(coldata))
)

# Regions we care about (biologically predefined sets)
regions <- GRanges(
  seqnames = "chr1",
  ranges   = IRanges(
    start = c(    1L, 10001L, 20001L, 30001L, 45001L),
    end   = c(10000L, 19000L, 28000L, 40000L, 55000L)
  ),
  feature_class = c("PRC", "CGI", "LAD", "PRC", "LAD")
)
names(regions) <- c("PRC_early", "CGI_island", "LAD_chr1q",
                    "PRC_mid",   "LAD_distal")

# Inject signal: in "old" samples, PRC_early gains ~1 M-value, LAD_chr1q loses ~1.
# CGI_island, PRC_mid, LAD_distal are left at baseline (should come out inconclusive).
old_mask <- coldata$group == "old"
probe_in <- function(gr) which(IRanges::overlapsAny(probes, gr))
X[probe_in(regions["PRC_early"]), old_mask] <-
  X[probe_in(regions["PRC_early"]), old_mask] + 1.0
X[probe_in(regions["LAD_chr1q"]), old_mask] <-
  X[probe_in(regions["LAD_chr1q"]), old_mask] - 1.0

se <- SummarizedExperiment(
  assays    = list(M = X),
  rowRanges = probes,
  colData   = coldata
)
se
```

## Fit BREAD

One call maps probes to regions, summarizes them per sample, fits a Bayesian
region-level model, and classifies each region as *hyper*, *hypo*, or
*inconclusive*.

```{r fit}
fit <- fit_bread(
  se,
  features    = regions,
  design      = ~ group + sex,
  contrast    = "groupold",
  assay_name  = "M",
  input_scale = "M",
  delta       = 0.10,
  prob_cutoff = 0.95,
  min_probes  = 3L,
  summary_fun = "mean",
  feature_class_col = "feature_class"
)
fit
```

The `<BreadFit>` printout summarizes the pipeline: mode, backend, contrast,
thresholds, how many of the supplied regions survived `min_probes`, and the
distribution of classifications.

## Inspect the region-level results

```{r results}
res <- results(fit)
res[, c("region_id", "mean_effect", "ci_lo", "ci_hi",
        "p_gt_delta", "p_lt_neg_delta", "classification")]
```

For a quick vector mapping region → call:

```{r classifications}
classifications(fit)
```

## Visualize

BREAD ships three plot helpers — all returning `ggplot` objects, all themed
with the embedded MetBrewer *Cross* palette via `bread_colors()`.

### 1. Posterior of the contrast coefficient

The analytical posterior of each region's `groupold` effect is a
location-scale Student-t. BREAD plots it with dashed guides at ±δ.

```{r plot-posterior, fig.height = 4}
plot_region_posterior(fit)
```

### 2. Raw region-level values by group

Region-level summaries plotted against the first design variable.

```{r plot-data, fig.width = 9, fig.height = 4}
p_hyper <- plot_region_data(fit, "PRC_early")
p_hypo  <- plot_region_data(fit, "LAD_chr1q")
p_hyper | p_hypo
```

### 3. Classification summary by feature class

```{r plot-feature-set, fig.height = 3.5}
plot_feature_set(fit, feature_class_col = "feature_class")
```

### Posterior samples

When you need draws (for custom plots, downstream uncertainty propagation,
or reporting), `posterior_draws()` samples from the analytical scaled-t:

```{r draws}
d <- posterior_draws(fit, region_id = "PRC_early", n = 2000L, seed = 1L)
head(d, 4)
quantile(d$value, c(0.025, 0.5, 0.975))
```

## What the model is doing (under the hood)

For each region *r*, BREAD (v1) fits a conjugate
Normal–Inverse-Gamma regression on the summarized region values:

\[
y_{ir} \sim \mathcal{N}\bigl(X_i \beta_r,\ \sigma_r^2\bigr),\quad
\beta_r \mid \sigma_r^2 \sim \mathcal{N}(\mu_0,\ \sigma_r^2 \Lambda_0^{-1}),\quad
\sigma_r^2 \sim \text{Inv-Gamma}(a_0, b_0).
\]

The posterior has a closed form. After the update, the **marginal** posterior
of any single coefficient is a location-scale Student-*t*:

\[
\beta_{r,k} \mid y \sim t_{2 a_n}\!\left(\mu_{n,k},\ \tfrac{b_n}{a_n} [\Lambda_n^{-1}]_{kk}\right),
\]

which is what BREAD uses to compute posterior probabilities directly via
`pt()` / `qt()` — no MCMC. This keeps v1 fast, deterministic, and trivially
installable. Partial pooling across regions and MCMC-backed priors are
planned for the hierarchical mode in v2.

Posterior probabilities that drive classification:

- \(P(\beta_{r,k} > \delta \mid y)\) → if \(\ge\) `prob_cutoff`, classify as **hypermethylated**
- \(P(\beta_{r,k} < -\delta \mid y)\) → if \(\ge\) `prob_cutoff`, classify as **hypomethylated**
- otherwise **inconclusive**.

You can customize the prior with `bread_prior()`:

```{r priors}
# A slightly tighter prior on coefficients
prior <- bread_prior(lambda0 = 0.1, a0 = 1, b0 = 1)
fit2  <- fit_bread(se, regions, ~ group, "groupold", prior = prior,
                   feature_class_col = "feature_class")
classifications(fit2)
```

## Session info

```{r session}
sessionInfo()
```
